### 2020-04-01 02-ReactionSystem.tex
major update
introdotto notazione, spostato flat, riveduto esempio 13

>>>>

Given an extended interactive process $\pi =(\gamma,\delta)$, we denote by $\pi^k$ the shift of $\pi$ starting at the $k$-th state sequence; formally we let $\pi^k=(\gamma^k,\delta^k)$ with
$\gamma^k=\{C_i'\}_{i\in\mathbb{N}}$, $\delta^k=\{D_i'\}_{i\in\mathbb{N}}$ with 
  $C_0'=C_k \cup D_k$, $D'_0 = \emptyset$, and   $C_i'=C_{i+k}$, $D_i'= D_{i+k}$ for any $i\geq 1$.

### 2020-04-01 05-translation.tex
major update
introdotto notazione, spostato flat, riveduto esempio 13

In the following use the following conventions for denoting different categories of names:
\begin{itemize}
\item
$\mathit{decs} \defeq \{ s, \overline{s}, \widetilde{s},\widehat{s},\underline{s}~|~ s\in S\}$ is the set of names for decorated entities;
\item
$\mathit{ents} \defeq \{ d_i, d_o~|~ d\in \mathit{decs}\}$ is the set names for entities;
\item
$\mathit{reacts}\defeq \{r_1,...,r_u\}$ is the set of reaction names associated with each reaction $aj$;
\item 
$\mathit{prods}\defeq \{p_1,...,p_u\}$ is the set of names for production $p_j$, associated with  reaction $aj$.
\end{itemize}

@@

 \begin{restatable}[]{lemma}{lemmastruct}
 \label{lem:struct}
%\begin{lemma}
 %\label{lem:struct}
  Let ${\cal A} = (S, A)$ be a RS and let $\pi=(\gamma,\delta)$ be an extended  interactive process in ${\cal A}$. Let $P=\llbracket{\cal A},\gamma\rrbracket$ its c\CNA~translation. If $exists$ $P'$ such that $t= (P \xrightarrow{\restrict{\mathit{names}}\upsilon} P')$ is a transition of $P$, then
  \begin{enumerate}
  \item for each reaction  $a_j \in A$, the corresponding channels $r_j$ and $p_j$ appear in  $\upsilon$; for each entity $s \in S$, the corresponding channel $s$ (suitably decorated) appear in $\upsilon$; 
  the channel $\mathit{cxt}$ appears in $\upsilon$;
  %processes $P_s$ (or $\overline{P_s}$) 
%  for each  $Cxt_s$ are
 %involved in the transition $t$;
  \item for each reaction $a \in A$ and each entity $s \in S$, each virtual link offered by processes $P_a$ and $\mathit{Cxt}^1$ is  overlapped by exactly one solid link offered by  processes representing entities.
  %each process $P_s $ (or its alternative version $\overline{P}_s$), with $s \in S$, participates at most  with one solid link for each process $P_a$, and with exactly one solid link for each process $Cxt_s$.%, if $s$ is a \emph{required element} of $A_j$ in $t$.

  \end{enumerate}
  \end{restatable}

\begin{example}\label{ex:backbone}
 Let ${\cal A}$ be a RS whose specification contains two entities, $s1$ and $s2$, and the reactions $r_1=(s1,\ , s2)$ and $r_2 = (s2, \ , s1)$ that produce $s2$ if $s1$ is present and $s1$ if $s2$ is present. For simplicity we do not consider inhibitors, but the reader can assume a void inhibitor is present in both reactions. Then, we assume an  extended interactive process $\pi=(\gamma,\delta)$ where the context $\gamma$ provides  $s1$ and $s2$ at every step, but we assume that only $s1$ is initially present. The corresponding c\CNA \ process is: 
$$\restrict{\mathit{names}} (~I ~|~ P_{s1}~|~\overline{P}_{s2}~|~ P_{r1}~|~P_{r2}~|~\mathit{Cxt})$$
 where:
 \[
 \begin{array}{l@{\hspace{-0.009cm}}c@{\hspace{-0.009cm}}ll@{\hspace{-0.009cm}}c@{\hspace{-0.009cm}}l}
 P_{r_1} &\defeq& \startchain{r_1}\chainedlink{s1_i}{\noact}\chainedlink{\noact}{s1_o}\chainedlink{r_2}{\noact} \chainedlink{\noact}{p_1}\chainedlink{\widetilde{s2}_i}{\noact}\chainedlink{\noact}{\widetilde{s2}_o}\chainend{p_2}.P_{r_1} + \dots; & 
   P_{s1} &\defeq& \startchain{s1_i}\chainedlink{s1_o}{\noact}\chainedlink{\noact}{\widehat{s1}_i}\chainend{\widehat{s1}_o}
 %{\noact}\chainedlink{\noact}{\tilde{s1}_i}\chainedlink{\tilde{s1}_o}
 .P_{s1} + \dots; \\   
  P_{r_2} &\defeq& \dots + \startchain{r_2}\chainedlink{\overline{s2}_i}{\noact}\chainedlink{\noact}{\overline{s2}_o}\chainedlink{cxt}{\noact} \chainedlink{\noact}{p_2}\chainend{\silent}.P_{r_2} + \dots; &
  \overline{P}_{s2} &\defeq& \startchain{\overline{s2}_i}\chainedlink{\overline{s2}_o}{\noact}
   \chainedlink{\noact} {\widetilde{s2}_i}\chainend{\widetilde{s2}_o}.P_{s2} + \dots;\\
 Cxt& \defeq & \startchain{cxt}\chainedlink{\widehat{s1}_i}{\noact}\chainedlink{\noact}{\widehat{s1}_o}
  \chainedlink{\widehat{s2}_i}{\noact}\chainedlink{\noact}{\widehat{s2}_o}\chainend{p_1}.Cxt\\

  %Cxt& \defeq & \startchain{cxt}\chainedlink{\hat{s1}_i}{\noact}\chainedlink{\noact}{\hat{s1}_o}
  %\chainedlink{\hat{s2}_i}{\noact}\chainedlink{\noact}{\hat{s2}_o}\chainend{p_1}.Cxt
 %Cxt_{s1}& \defeq & \startchain{cxt_1}\chainedlink{\hat{s1}_i}{\noact}\chainedlink{\noact}{\hat{s1}_o}\chainend{cxt_2}.Cxt_{s1} &
 % Cxt _{s2} &\defeq& \startchain{cxt_2}\chainedlink{\hat{s2}_i}{\noact}\chainedlink{\noact}{\hat{s2}_o}\chainend{p_1}. Cxt_{s2}
\end{array}
\]  \\
For clarity we show the code of the processes just in part, to focus on the prefixes that will be involved in the first transition of the system. In Figure~\ref{fig:backbone} we show the  structure of a link chain label related to the execution of such a transition. The yellow blocks are referred to init process $I$, to the processes encoding the reactions, $P_{r_1}$ and $P_{r_2}$, and to the context $\mathit{Cxt}$. As the figure puts in evidence, these two kinds of processes determine the structure of the link chain, from end to end, i.e. from the left $\silent$ to the right one. We could say that these processes form the \emph{backbone} of the interaction.
   In contrast, the processes encoding the entities, $P_{s1}$,  $\overline{P}_{s2}$, provides the
    solid links to be merged with the virtual links of the backbone (i.e. to be plugged in the backbone).

@@

With the next proposition, we analyse the structure of a c\CNA \ 
process encoding of  a reactive process after one step transition.
In the following four statements, for brevity, we let ${\cal A} = (S, A)$ be a RS, and let  $\pi=(\gamma,\delta)$ be an extended interactive process in $A$, with $\gamma=\{C_i\}_{i\in\mathbb{N}}$ and $\delta=\{D_i\}_{i\in\mathbb{N}}$. Moreover, we denote by $\pi^j$ the shift of $\pi$ starting at the $j$-th state sequence; formally we let $\pi^j=(\gamma^j,\delta^j)$ with
$\gamma^j=\{C_i'\}_{i\in\mathbb{N}}$, $\delta^j=\{D_i'\}_{i\in\mathbb{N}}$ with 
  $C_0'=C_j \cup D_j$, $D'_0 = \emptyset$, and   $C_i'=C_{i+j}$, $D_i'= D_{i+j}$ for any $i\geq 1$.

@@

\paragraph{The flat function}
For technical reasons, our transition labels are quite verbose; then, to simplify their processing, we introduce a function that takes a solid link chain and returns a simple string by eliminating all the channel matching pairs leaving just one channel per each pair. This transformation is harmless, in the sense that it retains all the information in the chain, because it is applied to solid chains only. Below we denote by $\mathit{decs} = \{ s, \overline{s}, \widetilde{s},\widehat{s},\underline{s}~|~ s\in S\}$ the set of decorated names (without subscripts $_i$ and $_o$).
The function $\mathit{flat}$ is defined inductively as follows:
%\[
%\begin{array}{lcl@{\hspace{0.5cm}}lcl@{\hspace{0.5cm}}lcl@{\hspace{0.5cm}}lcl}
%\mathit{flat}(\epsilon)&\defeq & \epsilon &
%\mathit{flat}(\link{\gamma}{\gamma'}) &  \defeq & \left \{ \begin{array}{cl} \epsilon & \mathrm{if } (\gamma = \beta_i \wedge \gamma = \beta_o)  \vee \gamma = \silent\\
%\gamma' & \mathrm{otw }  \end{array} \right. &
%%\mathit{flat}(\link{\alpha}{\silent})& \defeq & \epsilon &
%%\mathit{flat}(\link{\alpha}{\beta}) & \defeq &\beta \\
%\end{array}\]

\[
\mathit{flat}(\epsilon)\defeq \epsilon 
\qquad
\mathit{flat}(\link{\alpha}{\beta}) \defeq 
\left \{ \begin{array}{ll} 
\beta & \mbox{if $\beta\in \mathit{reacts}\cup\{\mathit{cxt}\}\cup\mathit{prods}$}\\
\gamma & \mbox{if $\beta=\gamma_i$ with $\gamma\in\mathit{decs}$}\\
\epsilon & \mbox{otherwise}  
\end{array} \right. 
\]

>>>>

In the following use the following conventions for denoting different categories of names:
\begin{itemize}
\item
$\mathit{decs} \defeq \{ s, \overline{s}, \widetilde{s},\widehat{s},\underline{s}~|~ s\in S\}$ is the set of names for decorated entities (without subscripts $_i$ and $_o$);
\item
$\mathit{ents} \defeq \{ d_i, d_o~|~ d\in \mathit{decs}\}$ is the set names for entities;
\item
$\mathit{reacts}\defeq \{r_1,...,r_u\}$ is the set of reaction names associated with each reaction $aj$;
\item 
$\mathit{prods}\defeq \{p_1,...,p_u\}$ is the set of names for production $p_j$, associated with  reaction $aj$.
\end{itemize}

@@

 \begin{restatable}[]{lemma}{lemmastruct}
 \label{lem:struct}
%\begin{lemma}
 %\label{lem:struct}
  Let ${\cal A} = (S, A)$ be a RS and let $\pi=(\gamma,\delta)$ be an extended  interactive process in ${\cal A}$. Let $P=\llbracket{\cal A},\gamma\rrbracket$ its c\CNA~translation. If $exists$ $P'$ such that $t= (P \xrightarrow{\restrict{\mathit{names}}\upsilon} P')$ is a transition of $P$, then
  \begin{enumerate}
  \item for each reaction  $a_j \in A$, the corresponding channels $r_j$ and $p_j$ appear in  $\upsilon$; for each entity $s \in S$, the corresponding channel $s$ (suitably decorated) appear in $\upsilon$; 
  the channel $\mathit{cxt}$ appears in $\upsilon$;
  %processes $P_s$ (or $\overline{P_s}$) 
%  for each  $Cxt_s$ are
 %involved in the transition $t$;
  \item for each reaction $a \in A$ and each entity $s \in S$, each virtual link offered by processes $P_a$ and $\mathit{Cxt}^1$ is  overlapped by exactly one solid link offered by  processes representing entities.
  %each process $P_s $ (or its alternative version $\overline{P}_s$), with $s \in S$, participates at most  with one solid link for each process $P_a$, and with exactly one solid link for each process $Cxt_s$.%, if $s$ is a \emph{required element} of $A_j$ in $t$.
  \end{enumerate}
  \end{restatable}

The topmost restriction $\restrict{\mathit{names}}$ appearing in the process $\llbracket{\cal A},\gamma\rrbracket$ serves to guarantee that all names appearing in a link of the chain labelling a transition are matched.
Since all names appearing in any prefix of $\llbracket{\cal A},\gamma\rrbracket$ are restricted, in the transition $\llbracket{\cal A},\gamma\rrbracket \xrightarrow{\restrict{\mathit{names}}\upsilon} P'$ it means that the observation $\restrict{\mathit{names}}\upsilon$ has the form $\link{\silent}{\silent}\dots\link{\silent}{\silent}$, i.e., it is silent, and that $\upsilon$ is solid. Later on we will be interested in reasoning about the actual chain $\upsilon$ used in the transition.  It has the peculiarity to start and end with silent actions and to include all names in $\mathit{reacts}\cup\mathit{prods}\cup\{\mathit{cxt}\}$. As a matter of notation we call such chain $\upsilon$ \emph{complete} (solid and with $\silent$ actions at the extremes). We will then use $\llparenthesis{\cal A},\gamma\rrparenthesis$ to refer to the encoding without topmost name restrictions, i.e.,
$$
\llparenthesis{\cal A},\gamma\rrparenthesis \defeq 
I~|~\Pi_{s\in C_0}P_{s} ~|~ \Pi_{s \notin C_0}\overline{P_s} ~|~ \Pi_{a\in A} P_a ~|~  \mathit{Cxt}^1
$$
and will write $P\xRightarrow{\upsilon} P'$ to mean that $P\xrightarrow{\upsilon} P'$ with $\upsilon$ complete.

\begin{example}\label{ex:backbone}
 Let ${\cal A}$ be a RS whose specification contains two entities, $s1$ and $s2$, and the reactions $r_1=(s1,\ , s2)$ and $r_2 = (s2, \ , s1)$ that produce $s2$ if $s1$ is present and $s1$ if $s2$ is present. For simplicity we do not consider inhibitors, but the reader can assume a void inhibitor is present in both reactions. Then, we assume an  extended interactive process $\pi=(\gamma,\delta)$ where the context $\gamma$ provides  $s1$ and $s2$ at every step, but we assume that only $s1$ is initially present. The corresponding c\CNA \ process is $\llbracket{\cal A},\gamma\rrbracket \defeq \restrict{\mathit{names}} \llparenthesis{\cal A},\gamma\rrparenthesis$, with 
$$
\llparenthesis{\cal A},\gamma\rrparenthesis \defeq 
I ~|~ P_{s1}~|~\overline{P}_{s2}~|~ P_{r1}~|~P_{r2}~|~\mathit{Cxt}
$$
 where:
 \[
 \begin{array}{l@{\hspace{-0.009cm}}c@{\hspace{-0.009cm}}ll@{\hspace{-0.009cm}}c@{\hspace{-0.009cm}}l}
 P_{r_1} &\defeq& \startchain{r_1}\chainedlink{s1_i}{\noact}\chainedlink{\noact}{s1_o}\chainedlink{r_2}{\noact} \chainedlink{\noact}{p_1}\chainedlink{\widetilde{s2}_i}{\noact}\chainedlink{\noact}{\widetilde{s2}_o}\chainend{p_2}.P_{r_1} + \dots; & 
   P_{s1} &\defeq& \startchain{s1_i}\chainedlink{s1_o}{\noact}\chainedlink{\noact}{\widehat{s1}_i}\chainend{\widehat{s1}_o}
 %{\noact}\chainedlink{\noact}{\tilde{s1}_i}\chainedlink{\tilde{s1}_o}
 .P_{s1} + \dots; \\   
  P_{r_2} &\defeq& \dots + \startchain{r_2}\chainedlink{\overline{s2}_i}{\noact}\chainedlink{\noact}{\overline{s2}_o}\chainedlink{\mathit{cxt}}{\noact} \chainedlink{\noact}{p_2}\chainend{\silent}.P_{r_2} + \dots; &
  \overline{P}_{s2} &\defeq& \startchain{\overline{s2}_i}\chainedlink{\overline{s2}_o}{\noact}
   \chainedlink{\noact} {\widehat{s2}_i}\chainedlink{\widehat{s2}_o}{\noact}\chainedlink{\noact} {\widetilde{s2}_i}\chainend{\widetilde{s2}_o}.P_{s2} + \dots;\\
 \mathit{Cxt}& \defeq & \startchain{\mathit{cxt}}\chainedlink{\widehat{s1}_i}{\noact}\chainedlink{\noact}{\widehat{s1}_o}
  \chainedlink{\widehat{s2}_i}{\noact}\chainedlink{\noact}{\widehat{s2}_o}\chainend{p_1}.\mathit{Cxt}
  %Cxt& \defeq & \startchain{cxt}\chainedlink{\hat{s1}_i}{\noact}\chainedlink{\noact}{\hat{s1}_o}
  %\chainedlink{\hat{s2}_i}{\noact}\chainedlink{\noact}{\hat{s2}_o}\chainend{p_1}.Cxt
 %Cxt_{s1}& \defeq & \startchain{cxt_1}\chainedlink{\hat{s1}_i}{\noact}\chainedlink{\noact}{\hat{s1}_o}\chainend{cxt_2}.Cxt_{s1} &
 % Cxt _{s2} &\defeq& \startchain{cxt_2}\chainedlink{\hat{s2}_i}{\noact}\chainedlink{\noact}{\hat{s2}_o}\chainend{p_1}. Cxt_{s2}
\end{array}
\]  \\
For clarity of exposition, we show the code of the processes just in part, to focus on the prefixes that will be involved in the first transition of the system. In Figure~\ref{fig:backbone} we show the  structure of a link chain label related to the execution of such a transition. The yellow blocks are referred to init process $I$, to the processes encoding the reactions, $P_{r_1}$ and $P_{r_2}$, and to the context $\mathit{Cxt}$. As the figure puts in evidence, these two kinds of processes determine the structure of the link chain, from end to end, i.e. from the left $\silent$ to the right one. We could say that these processes form the \emph{backbone} of the interaction.
   In contrast, the processes encoding the entities, $P_{s1}$,  $\overline{P}_{s2}$, provides the
    solid links to be merged with the virtual links of the backbone (i.e. to be plugged in the backbone).
 Formally, we have
 $$
 \llparenthesis{\cal A},\gamma\rrparenthesis \xRightarrow{
 \startchain{\silent}
 \chainedlink{r_1}{r_1}
 \chainedlink{s1_i}{s1_i}
 \chainedlink{s1_o}{s1_o}
 \chainedlink{r_2}{r_2}
 \chainedlink{\overline{s2}_i}{\overline{s2}_i}
 \chainedlink{\overline{s2}_o}{\overline{s2}_o}
 \chainedlink{\mathit{cxt}}{\mathit{cxt}}
 \chainedlink{\widehat{s1}_i}{\widehat{s1}_i}
 \chainedlink{\widehat{s1}_o}{\widehat{s1}_o}
  \chainedlink{\widehat{s2}_i}{\widehat{s2}_i}
  \chainedlink{\widehat{s2}_o}{\widehat{s2}_o}
  \chainedlink{p_1}{p_1}
  \chainedlink{\widetilde{s2}_i}{\widetilde{s2}_i}
  \chainedlink{\widetilde{s2}_o}{\widetilde{s2}_o}
  \chainedlink{p_2}{p_2}
  \chainend{\silent}
 }
 P'
 $$

@@

\paragraph{The flat function}
For technical reasons, our transition labels are quite verbose; then, to simplify their processing, we introduce a function that takes a solid link chain and returns a simple string by eliminating all the channel matching pairs leaving just one placeholder for them. This transformation is harmless, in the sense that it retains all the information in the chain, because it is applied to complete chains only. 
The function $\mathit{flat}$ is defined inductively as follows:
%\[
%\begin{array}{lcl@{\hspace{0.5cm}}lcl@{\hspace{0.5cm}}lcl@{\hspace{0.5cm}}lcl}
%\mathit{flat}(\epsilon)&\defeq & \epsilon &
%\mathit{flat}(\link{\gamma}{\gamma'}) &  \defeq & \left \{ \begin{array}{cl} \epsilon & \mathrm{if } (\gamma = \beta_i \wedge \gamma = \beta_o)  \vee \gamma = \silent\\
%\gamma' & \mathrm{otw }  \end{array} \right. &
%%\mathit{flat}(\link{\alpha}{\silent})& \defeq & \epsilon &
%%\mathit{flat}(\link{\alpha}{\beta}) & \defeq &\beta \\
%\end{array}\]

\[
\mathit{flat}(\epsilon)\defeq \epsilon 
\qquad
\mathit{flat}(\link{\alpha}{\beta}) \defeq 
\left \{ \begin{array}{ll} 
\beta & \mbox{if $\beta\in \mathit{reacts}\cup\{\mathit{cxt}\}\cup\mathit{prods}$}\\
d & \mbox{if $\beta=d_i$ with $d\in\mathit{decs}$}\\
\epsilon & \mbox{otherwise}  
\end{array} \right. 
\]

For example, if we consider again the complete label
$$
\upsilon =  \startchain{\silent}
 \chainedlink{r_1}{r_1}
 \chainedlink{s1_i}{s1_i}
 \chainedlink{s1_o}{s1_o}
 \chainedlink{r_2}{r_2}
 \chainedlink{\overline{s2}_i}{\overline{s2}_i}
 \chainedlink{\overline{s2}_o}{\overline{s2}_o}
 \chainedlink{\mathit{cxt}}{\mathit{cxt}}
 \chainedlink{\widehat{s1}_i}{\widehat{s1}_i}
 \chainedlink{\widehat{s1}_o}{\widehat{s1}_o}
  \chainedlink{\widehat{s2}_i}{\widehat{s2}_i}
  \chainedlink{\widehat{s2}_o}{\widehat{s2}_o}
  \chainedlink{p_1}{p_1}
  \chainedlink{\widetilde{s2}_i}{\widetilde{s2}_i}
  \chainedlink{\widetilde{s2}_o}{\widetilde{s2}_o}
  \chainedlink{p_2}{p_2}
  \chainend{\silent}
$$
from Example~\ref{ex:backbone}, we have
$$
\mathit{flat}(\upsilon) = 
r_1\ 
s1\ 
r_2\
\overline{s2}\ 
\mathit{cxt}\ 
\widehat{s1}\ 
\widehat{s2}\ 
p_1\ 
\widetilde{s2}\ 
p_2.
$$

Viceversa, it is immediate to define the function $\mathit{unflat}$ to rebuild the complete label from the compact string (here we exploit again the half link and block notation):
$$
\mathit{unflat}(x) \defeq\left\{
\begin{array}{ll}
^{x}_{x} & \mbox{if $x\in \mathit{reacts}\cup\{\mathit{cxt}\}\cup\mathit{prods}$}\\
^{x_i}_{x_i}\backslash^{x_o}_{x_o} & \mbox{if $x\in \mathit{decs}$}\\
\end{array}
\right.
$$
$$
\mathit{unflat}(x_1\dots x_n) \defeq\ ^{\silent}\backslash \mathit{unflat}(x_1) \backslash \dots \backslash \mathit{unflat}(x_1) \backslash_{\silent}
$$

It is immediate to check that for any complete label $\upsilon$ of our processes we have
$\upsilon = \mathit{unflat}(\mathit{flat}(\upsilon))$.

With the next proposition, we analyse the structure of a c\CNA \ 
process encoding of  a reactive process after one step transition.
In the following four statements, for brevity, we let ${\cal A} = (S, A)$ be a RS, and let  $\pi=(\gamma,\delta)$ be an extended interactive process in $A$, with $\gamma=\{C_i\}_{i\in\mathbb{N}}$ and $\delta=\{D_i\}_{i\in\mathbb{N}}$.

@@
